Comparison of two object-oriented languages: Eiffel and Ruby

### 2014-01-13 paper.tex
metamodel

This is a feature that improves the reliability of the produced software and shows again that Eiffel's primary concern is to enhance the software quality. In Ruby there is nothing like void-safety. Void safety is achieved at compile time and Ruby is an interpreted language, thus there is no way to make it void safe except by checking if the object is the nil object.
\subsection{Ruby}
Ruby is a programming language known for being very object-oriented and for its simplicity. It is also very flexible like duck-typing already attested in previous section.
\subsubsection{Open Classes}
In Ruby it is always possible to add new methods to an existing class, the class definitions can at any moment be opened for modifications. Even built-in classes are can be adapted. Adding or modifying content at run-time to an already existing class definition without altering the source code is known as \emph{monkey patching} \cite{wiki:monkey}. 
\\
\\
Monkey patching can be useful for different applications like extending or modifying the behaviour of an object at run-time from a third-party software without changing the source code. This enables for example to reuse existing code and adapt it in different files with different behaviours.
\\
\\
In first instance this feature seems really helpful because it gives the possibility to the developer to adapt code when needed. For example, as workaround to a bug by implementing a specific solution in this part of the code. But with great power comes great responsibility. Monkey patching can lead to bugs that break the code. If the source code changes, it is possible that the patch behaves differently because it makes assumptions on the code that do not hold anymore. Different patches for the same method create conflicts and one patch will override the other. Thus it is important to be aware of the patches and save the implementation of a method if necessary by renaming it with an alias. It is also important to document the patches and method that are patched, this can prevent confusion if one forgets or does not know about the existing patches and does not understand why the code does not act like expected.
\\
\\
To summarize, open classes is thus a language feature that offers a lot of modularity and flexibility. It enable to extend and reuse third-party code in a simple way. This is certainly why this feature has been added in Ruby. But it is very important to note that open classes are in conflict with the principle of encapsulation. Every method can be accessed with open classes, even private methods and they can  be changed into public ones. It is thus really important of not abusing of this feature. 
\subsubsection{Metaclass model}
\label{sec:metaclass}
% http://en.wikipedia.org/wiki/Metaclass
A metaclass is a class that has classes as instances in object-oriented context. It defines how a class and its instances have to work.

\section{Conclusion}
It is clear that even if Eiffel and Ruby are both object-oriented languages, they are very different in many aspects.
\section{References}
\bibliographystyle{plain}
\bibliography{paper}
\newpage
\section{Code listings}
% ACCESS CODE
\lstinputlisting[label=rb:access,language=Ruby,caption={Access Control in Ruby}]{code_Ruby/access.rb}
\lstinputlisting[label=e:access,language=Eiffel,caption={Access Control in Eiffel}]{code_Eiffel/access.e}

@@

\end{document}

>>>>

This is a feature that improves the reliability of the produced software and shows again that Eiffel's primary concern is to enhance the software quality. In Ruby there is nothing like void-safety. Void safety is achieved at compile time and Ruby is an interpreted language, thus there is no way to make it void safe except by checking if the object is the nil object.
\subsection{Ruby}
Ruby is a programming language known for being very object-oriented and for its simplicity. It is also very flexible like duck-typing already attested in previous section.
\subsubsection{Open Classes}
In Ruby it is always possible to add new methods to an existing class, the class definitions can at any moment be opened for modifications. Even built-in classes are can be adapted. Adding or modifying content at run-time to an already existing class definition without altering the source code is known as \emph{monkey patching} \cite{wiki:monkey}. 
Monkey patching can be useful for different applications like extending or modifying the behaviour of an object at run-time from a third-party software without changing the source code. This enables for example to reuse existing code and adapt it in different files with different behaviours.
In first instance this feature seems really helpful because it gives the possibility to the developer to adapt code when needed. For example, as workaround to a bug by implementing a specific solution in this part of the code. But with great power comes great responsibility. Monkey patching can lead to bugs that break the code. If the source code changes, it is possible that the patch behaves differently because it makes assumptions on the code that do not hold anymore. Different patches for the same method create conflicts and one patch will override the other. Thus it is important to be aware of the patches and save the implementation of a method if necessary by renaming it with an alias. It is also important to document the patches and method that are patched, this can prevent confusion if one forgets or does not know about the existing patches and does not understand why the code does not act like expected.
To summarize, open classes is thus a language feature that offers a lot of modularity and flexibility. It enable to extend and reuse third-party code in a simple way. This is certainly why this feature has been added in Ruby. But it is very important to note that open classes are in conflict with the principle of encapsulation. Every method can be accessed with open classes, even private methods and they can  be changed into public ones. It is thus really important of not abusing of this feature. 
\subsubsection{Metaclass model}
\label{sec:metaclass}
% http://en.wikipedia.org/wiki/Metaclass
A metaclass is a special class that has classes as instances. It defines thus methods that every class and its instances should have. When asking to a class in Ruby from which class it is an instance, it will always return the class Class. This is logical since every class should inherit the methods defined in it. Thus the only metaclass in Ruby is thus the class Class.
But lets look the meta-system a little bit more in details. In Ruby there exist meta-objects called \emph{eigenclasses}. Every object has its own eigenclass. Because classes are also objects in Ruby, they also have eigenclasses. And what is interesting is that eigenclasses also have eigenclasses. This is recursive and goes on forever. Thus there are three sorts of eigenclasses: eigenclasses for objects and for classes. 
But what is the usefulness of theses eigenclasses? When an instance method is defined, it is stored in the class. When a class method is defined, it is stored in the class' eigenclass. Thus the eigenclass serves to store class variables and class methods. Eigenclasses of objects are useful if the developer wants that this particular object has a different behaviour than the other objects of the same class. Specialized method for this object will be stored in the eigenclass of the object. And finally, Eigenclasses of eigenclasses do not store any method or data, thus the reason of there existence is for conceptually correctness. 
\includegraphics[scale=0.5]{Ruby.png}
\caption{The Ruby metaclass model. Image from Wikipedia}
\label{fig:metaclass}
The method lookup mechanism works thus as follows: 
\begin{enumerate}
	\item A method is called on the object
	\item The method is looked up in the methods of the object's eigenclass
	\item If the method is not found, the lookup continues in the class of the object
	\item If the method is still not found, it looks the whole inheritance structure up to end with the class Class. 
\end{enumerate}
The same lookup procedure is applied for class method calls, but then it is the eigenclass class that is looked up. Also it is interesting to note that the superclass of the eigenclass of the BasicObject class is the class Class, while the superclass of the BasicObject class is nil.
This metamodel description gives more explanation on how everything in Ruby are object. It also emphasizes another way of how Ruby can be flexible, every object can have its own specialized behaviour if the behaviour is added to its eigenclass object. Eiffel has no such metaclass model, even if everything is an object. 
\section{Conclusion}
It is clear that even if Eiffel and Ruby are both object-oriented languages, they are very different in many aspects. Eiffel tries to achieve reliability through controlling everything. A good example is the selective export. It defines everything that may be exported and specifies a detailed list of clients. Ruby is more into the simplicity and flexibility. It gives the developer the tools for doing whatever they want and for complete different programming paradigms. The philosophies of the two languages are thus really different. Eiffel would probably be more used in the industry where people need a good framework for producing quality software. Ruby is a higher level programming language that can be used for a lot of different purposes, going from small to larger project. But in large project Ruby seems less adequate than Eiffel. Developers need to be aware of possibilities of Ruby's flexibility and may not abuse of it, otherwise the software will be difficult to maintain.
\section{References}
\bibliographystyle{plain}
\bibliography{paper}
\newpage
\section{Code listings}
% ACCESS CODE
\lstinputlisting[label=rb:access,language=Ruby,caption={Access Control in Ruby}]{code_Ruby/access.rb}
\lstinputlisting[label=e:access,language=Eiffel,caption={Access Control in Eiffel}]{code_Eiffel/access.e}

@@

% METACLASS CODE
\lstinputlisting[label=rb:metaclass,language=Ruby,caption={Metamodel in Ruby}]{code_Ruby/metaclass.rb}
\end{document}
